---
title: "Prediction Of Manner Of The Exercise"
author: "Swapnil Jamthe"
date: "19 June 2015"
output: html_document
---
##Summary  
Data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants was collected. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Models are built to predict the manner in which they did the exercise.
More information is available at  http://groupware.les.inf.puc-rio.br/har.

##Shortlisting features  
Removed following columns from the data.  
1. Columns with more than 50% of values having "NA".  
2. Columns with more than 50% of empty values.  
3. user_name, new_window, num_window columns.  

##Dividing data  
Data is divided into following.  
1. 20% of data is used for calculation of out of sample error.  
2. Out of remaining data 20% is used as cross validation of different models.  
3. Remaining data is used as training set.   

##Prediction Models  
Following prediction models are tried which are trained on training set.  
1. rpart5 - Classification tree with 5 fold cross validation  
2. rpart10 - Classification tree with 10 fold cross validation  
3. rf5 - Random forest with 5 fold cross validation  
4. gbm5 - Boosted trees with 5 fold cross validation  
5. gbm10 - Boosted tress with 10 fold cross validation  


```{r echo=FALSE,message=FALSE,include=FALSE}
library(caret)
library(doParallel)
activityData<-read.csv("pml-training.csv")
naCols<-sapply(activityData,function(x){ifelse((sum(is.na(x))/nrow(activityData))<0.5,TRUE,FALSE)})
activityDataNAFilter<-activityData[,naCols];
emptyCols<-sapply(activityDataNAFilter,function(x){ifelse((sum(x=="")/nrow(activityData))<0.5,TRUE,FALSE)})
activityDataEmptyFilter<-activityDataNAFilter[,emptyCols]
timeCols<-sapply(names(activityDataEmptyFilter),function(x){ifelse(grepl("timestamp",x),FALSE,TRUE)})
activityDataFinal<-activityDataEmptyFilter[,timeCols]
activityDataFinal$user_name<-NULL
activityDataFinal$new_window<-NULL
activityDataFinal$num_window<-NULL
activityDataFinal$X<-NULL
trainingIndex<-createDataPartition(y=activityDataFinal$classe,p=0.8,list=FALSE)
trainingData<-activityDataFinal[trainingIndex,]
testingData<-activityDataFinal[-trainingIndex,]
trainingIndex<-createDataPartition(y=trainingData$classe,p=0.8,list=FALSE)
trainingData<-trainingData[trainingIndex,]
validationData<-trainingData[-trainingIndex,]
```



```{r echo=FALSE,message=FALSE,cache=TRUE,include=FALSE}
cl <- makeCluster(detectCores())
registerDoParallel(cl)
tc5 <- trainControl("cv",5)
tc10 <- trainControl("cv",10)
modFitRpart5<-train(classe~.,data = trainingData,trControl=tc5,method="rpart")
modFitRpart10<-train(classe~.,data = trainingData,trControl=tc10,method="rpart")
modFitRf5<-train(classe~.,data = trainingData,trControl=tc5,method="rf")
modFitGbm5<-train(classe~.,data = trainingData,trControl=tc5,method="gbm")
modFitGbm10<-train(classe~.,data = trainingData,trControl=tc10,method="gbm")
stopCluster(cl)
```

```{r echo=FALSE,message=FALSE}
results<-resamples(list(rpart5=modFitRpart5,rf5=modFitRf5,gbm5=modFitGbm5))
bwplot(results,main="Comparison 5 fold cross validation models")
```

Box plot shows random forest model has better performance than other two.

```{r echo=FALSE,message=FALSE,include=FALSE}
predRPart5<-predict(modFitRpart5,newdata=validationData)
rpart5Acc<-sum(predRPart5 == validationData$classe)/nrow(validationData)
predRPart10<-predict(modFitRpart10,newdata=validationData)
rpart10Acc<-sum(predRPart10 == validationData$classe)/nrow(validationData)
predRf5<-predict(modFitRf5,newdata=validationData)
rf5Acc<-sum(predRf5 == validationData$classe)/nrow(validationData)
predGbm5<-predict(modFitGbm5,newdata=validationData)
gbm5Acc<-sum(predGbm5 == validationData$classe)/nrow(validationData)
predGbm10<-predict(modFitGbm10,newdata=validationData)
gbm10Acc<-sum(predGbm10 == validationData$classe)/nrow(validationData)
```

Following are the accuracies of the models on cross validation set for models.  
1. rpart5 - `r rpart5Acc`  
2. rpart10 - `r rpart10Acc`  
3. rf5 - `r rf5Acc`  
4. gbm5 - `r gbm5Acc`  
5. gbm10 - `r gbm10Acc`  

```{r echo=FALSE,message=FALSE,include=FALSE}
predRf5OOS<-predict(modFitRf5,newdata=testingData)
rf5AccOOS<-sum(predRf5OOS == testingData$classe)/nrow(testingData)
oose<-1 - rf5AccOOS

```


##Conclusion  
Selecting model gbm5 (Random forest with 5 fold cross validation) which has highest accuracy on cross validation set for models. Out of sample error is `r oose`.





